Titeln på avhandlingens språk

### 2022-12-19 body/glossary.tex
glossaries

\newabbreviation{LAN}{LAN}{Local-Area Network}
\newabbreviation{WLAN}{WLAN}{Wireless Local-Area Network}
\newabbreviation{KPI}{KPI}{Key Performance Indicator}
\newabbreviation{WPAN}{WPAN}{Wireless Personal Area Network}
\newabbreviation{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArkin`g` serVice on the Edge}
\newabbreviation{OS}{OS}{Operating System}
\newabbreviation{UDP}{UDP}{User Datagram Protocol}
\newabbreviation{TCP}{TCP}{Transmission Control Protocol}
\newabbreviation{RMS}{RMS}{Root Mean Square}
\newabbreviation{RTT}{RTT}{Round-Trip Time}
\newabbreviation{CI}{CI}{Confidence Interval}
\newabbreviation{AP}{AP}{Access Point}
\newabbreviation{API}{API}{Application Programming Interface}
\newabbreviation{SSF}{SSF}{Swedish Foundation for Strategic Research}
\newabbreviation{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
\newabbreviation{ABC}{ABC}{Abstract Base Class}
%\acroplural{ABC}[ABCs]{Abstract Base Classes}

\newabbreviation{URL}{URL}{Uniform Record Locator}
\newabbreviation{AWS}{AWS}{Amazon Web Services}
\newabbreviation{EC2}{EC2}{Elastic Compute 2}
\newabbreviation{AMI}{AMI}{Amazon Machine Image}
\newabbreviation{SSH}{SSH}{Secure Shell}
\newabbreviation{IP}{IP}{Internet Protocol}
\newabbreviation{VPN}{VPN}{Virtual Private Network}
\newabbreviation{NAT}{NAT}{Network Address Translation}
\newabbreviation{NTP}{NTP}{Network Time Protocol}
%\acroindefinite{NTP}{an}{a}

\newabbreviation{SDR}{SDR}{Software-Defined Radio}
%\acroindefinite{SDR}{an}{a}

\newabbreviation{VLAN}{VLAN}{Virtual Local-Area Network}
\newabbreviation{APT}{APT}{Advaced Packaging Tool}
\newabbreviation{LTE}{LTE}{Long-Term Evolution}
\newabbreviation{SSID}{SSID}{Service Set Identifier}
\newabbreviation{EPC}{EPC}{Evolved Packet Core}
\newabbreviation{UE}{UE}{User Equipment}
\newabbreviation{eNodeB}{eNodeB}{Evolved Node B}
\newabbreviation{DNS}{DNS}{Domain Name System}
\newabbreviation{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
\newabbreviation{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
%\acroindefinite{HTTP}{an}{a}

\newabbreviation{YAML}{YAML}{YAML Ain't Markup Language}
\newabbreviation{O-RAN}{O-RAN}{Open Radio Access Network}
\newabbreviation{FOSS}{FOSS}{Free and Open Source Software}
\newabbreviation{COSMOS}{COSMOS}{Cloud enhanced Open Software defined MObile wireless testbed for city-Scale deployment}
\newabbreviation{OMF}{OMF}{ORBIT Management Framework}
\newabbreviation{POWDER}{POWDER}{Platform for Open Wireless Data-driven Experimental Research}
\newabbreviation{COTS}{COTS}{Commercial Off-The-Shelves}
\newabbreviation{RF}{RF}{Radio-Frequency}
\newabbreviation{VM}{VM}{Virtual Machine}
\newabbreviation{REST}{REST}{REpresentational Sate Transfer}
\newabbreviation{MaaS}{MaaS}{Metal-as-a-Service}
\newabbreviation{OEDL}{OEDL}{\ac{OMF} Experiment Description Language}
\newabbreviation{RSpec}{RSpec}{Resource Specification}
\newabbreviation{LXC}{LXC}{LinuX Containers}
\newabbreviation{MEC}{MEC}{Mobile Edge Computing}
\newabbreviation{USRP}{USRP}{Universal Software Radio Peripheral}
\newabbreviation{OAI}{OAI}{OpenAirInterface}
\newabbreviation{NR}{NR}{New Radio}
\newabbreviation{FPGA}{FPGA}{Field-Programmable Gate Array}
\newabbreviation{ARA}{ARA}{Agricultural and ruRAl communities}
\newabbreviation{UHF}{UHF}{Ultra-High Frequency}
\newabbreviation{MIMO}{MIMO}{Multi-Input Multi-Output}
\newabbreviation{CHI}{CHI}{CHameleon Infrastructure}
\newabbreviation{CMU}{CMU}{Carnegie Mellon University}
\newabbreviation{GSR}{GSR}{Galvanic Skin Response}
\newabbreviation{EEG}{EEG}{electroencephalography}
\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits~\cite{john1999big}}
\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}

>>>>

\newabbreviation{LAN}{LAN}{Local-Area Network}
\newabbreviation{WLAN}{WLAN}{Wireless Local-Area Network}
\newabbreviation{KPI}{KPI}{Key Performance Indicator}
\newabbreviation{WPAN}{WPAN}{Wireless Personal Area Network}
\newabbreviation{CSMA/CD}{CSMA/CD}{Carrier-Sense Multiple Access with Collision Detection}
\newabbreviation{OS}{OS}{Operating System}
\newabbreviation{UDP}{UDP}{User Datagram Protocol}
\newabbreviation{TCP}{TCP}{Transmission Control Protocol}
\newabbreviation{RMS}{RMS}{Root Mean Square}
\newabbreviation{RTT}{RTT}{Round-Trip Time}
\newabbreviation{CI}{CI}{Confidence Interval}
\newabbreviation{AP}{AP}{Access Point}
\newabbreviation{API}{API}{Application Programming Interface}
\newabbreviation{SSF}{SSF}{Swedish Foundation for Strategic Research}
\newabbreviation{TECoSA}{TECoSA}{Trustworthy Edge Computing Systems and Applications}
\newabbreviation{ABC}{ABC}{Abstract Base Class}
%\acroplural{ABC}[ABCs]{Abstract Base Classes}
\newabbreviation{URL}{URL}{Uniform Record Locator}
\newabbreviation{AWS}{AWS}{Amazon Web Services}
\newabbreviation{EC2}{EC2}{Elastic Compute 2}
\newabbreviation{AMI}{AMI}{Amazon Machine Image}
\newabbreviation{SSH}{SSH}{Secure Shell}
\newabbreviation{IP}{IP}{Internet Protocol}
\newabbreviation{VPN}{VPN}{Virtual Private Network}
\newabbreviation{NAT}{NAT}{Network Address Translation}
\newabbreviation{NTP}{NTP}{Network Time Protocol}
%\acroindefinite{NTP}{an}{a}
\newabbreviation{SDR}{SDR}{Software-Defined Radio}
%\acroindefinite{SDR}{an}{a}
\newabbreviation{VLAN}{VLAN}{Virtual Local-Area Network}
\newabbreviation{APT}{APT}{Advaced Packaging Tool}
\newabbreviation{LTE}{LTE}{Long-Term Evolution}
\newabbreviation{SSID}{SSID}{Service Set Identifier}
\newabbreviation{EPC}{EPC}{Evolved Packet Core}
\newabbreviation{UE}{UE}{User Equipment}
\newabbreviation{eNodeB}{eNodeB}{Evolved Node B}
\newabbreviation{DNS}{DNS}{Domain Name System}
\newabbreviation{MNIST}{MNIST}{Modified National Institute of Standards and Technology}
\newabbreviation{HTTP}{HTTP}{Hyper-Text Transfer Protocol}
%\acroindefinite{HTTP}{an}{a}
\newabbreviation{YAML}{YAML}{YAML Ain't Markup Language}
\newabbreviation{O-RAN}{O-RAN}{Open Radio Access Network}
\newabbreviation{FOSS}{FOSS}{Free and Open Source Software}
\newabbreviation{COSMOS}{COSMOS}{Cloud enhanced Open Software defined MObile wireless testbed for city-Scale deployment}
\newabbreviation{OMF}{OMF}{ORBIT Management Framework}
\newabbreviation{POWDER}{POWDER}{Platform for Open Wireless Data-driven Experimental Research}
\newabbreviation{COTS}{COTS}{Commercial Off-The-Shelves}
\newabbreviation{RF}{RF}{Radio-Frequency}
\newabbreviation{VM}{VM}{Virtual Machine}
\newabbreviation{REST}{REST}{REpresentational Sate Transfer}
\newabbreviation{MaaS}{MaaS}{Metal-as-a-Service}
\newabbreviation{OEDL}{OEDL}{\gls{OMF} Experiment Description Language}
\newabbreviation{RSpec}{RSpec}{Resource Specification}
\newabbreviation{LXC}{LXC}{LinuX Containers}
\newabbreviation{MEC}{MEC}{Mobile Edge Computing}
\newabbreviation{USRP}{USRP}{Universal Software Radio Peripheral}
\newabbreviation{OAI}{OAI}{OpenAirInterface}
\newabbreviation{NR}{NR}{New Radio}
\newabbreviation{FPGA}{FPGA}{Field-Programmable Gate Array}
\newabbreviation{ARA}{ARA}{Agricultural and ruRAl communities}
\newabbreviation{UHF}{UHF}{Ultra-High Frequency}
\newabbreviation{MIMO}{MIMO}{Multi-Input Multi-Output}
\newabbreviation{CHI}{CHI}{CHameleon Infrastructure}
\newabbreviation{CMU}{CMU}{Carnegie Mellon University}
\newabbreviation{GSR}{GSR}{Galvanic Skin Response}
\newabbreviation{EEG}{EEG}{electroencephalography}
\newabbreviation{BFI}{BFI}{Big Five Inventory of personality traits~\cite{john1999big}}
\newabbreviation{ITQ}{ITQ}{Immersive Tendencies Questionnaire~\cite{witmer1998measuring}}
\newabbreviation{GPS}{GPS}{Global Positioning System}
\newabbreviation{FSM}{FSM}{Finite State Machine}
%\acroindefinite{NCS}{an}{a}
\newabbreviation{CLEAVE}{CLEAVE}{ControL bEnchmArking serVice on the Edge}
\newabbreviation{TTF}{TTF}{Time-to-Feedback}
\newabbreviation{MAR}{MAR}{Mobile \gls{AR}}
\newabbreviation{QoE}{QoE}{Quality of Experience}
\newabbreviation{QoS}{QoS}{Quality of Service}
\newabbreviation{PCA}{PCA}{Principal Component Analysis}
\newabbreviation{exGaussian}{Ex-Gaussian}{Exponentially Modified Gaussian}
\newabbreviation{IEEE}{IEEE}{Institute of Electrical and Electronics Engineers}
\newabbreviation{CPU}{CPU}{Central Processing Unit}
\newabbreviation{N/A}{N/A}{Not Applicable}
\newabbreviation{CIO}{CIO}{Confidence Interval}
\newabbreviation{MLE}{MLE}{Maximum Likelihood Estimation}
\newabbreviation{CDF}{CDF}{Cumulative Distribution Function}
\newabbreviation{CCDF}{CCDF}{Complementary \gls{CDF}}
\newabbreviation{PDF}{PDF}{Probability Density Function}
\newabbreviation{ECDF}{ECDF}{Empirical \gls{CDF}}
\newabbreviation{NSF}{NSF}{United States National Science Foundation}

### 2022-12-19 body/part1/introduction.tex
glossaries

This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \ac{WCA} and \ac{NCS}.
This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
We validate this methodology by example, presenting implementations of its application in the context of \ac{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.

Secondly, we present a deeper exploration of this methodology for \ac{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \ac{WCA}.
This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.

Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \ac{WCA} deployments.
In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
We conclude that significant improvements can be achieved through the use of our human user model.

@@

\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \ac{WCA}, making very straightforward assumptions about human behavior in \ac{WCA} applications.
\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.

\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \acp{NCS}.
As discussed above, these applications are similarly difficult as \ac{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
\todo[inline]{More argumentation or refer to previous section.}
We apply thus in this work our methodology to \ac{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
The software acts as a middleware which abstracts away the network from the development of these workloads, allowing for quick prototyping and deployment.
This work also showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.

@@

Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \ac{WCA}, as well as study its implications for optimization of these systems.
Although the above approach to human behavior in \ac{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \ac{WCA} application.
System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.

\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \ac{WCA}.
The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
This work also explores potential implications of this model for \ac{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.

>>>>

This thesis presents three core contributions to the existing body of research in edge computing, as well as a number of secondary ones.
Firstly, we introduce a methodological approach to studying system responsiveness versus resource consumption trade-offs in edge-bound feedback-loop systems such as \gls{WCA} and \gls{NCS}.
This approach is based on the emulation of the client-side workload, while maintaining the \emph{real} server-side process as well as network stack.
We validate this methodology by example, presenting implementations of its application in the context of \gls{WCA} and \acp{NCS}, and introduce a software framework for the orchestration of the edge computing testbeds necessary for its implementation.

Secondly, we present a deeper exploration of this methodology for \gls{WCA} and introduce the first ever model for the end-to-end emulation of human timing behavior in \gls{WCA}.
This model is built from a thorough characterization of these behaviors, the data for which we obtain from a comprehensive human-subject study.

Finally, as an extension and application of our first and second contributions, we explore the implications of our methodology and human user model on the optimization potential of \gls{WCA} deployments.
In particular, we study the sampling and energy consumption behaviors of these systems with and without considering realistic human behavior.
We conclude that significant improvements can be achieved through the use of our human user model.

@@

\cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid,paper:olguinmunoz2022cleave,paper:olguinmunoz2022ainur} introduce and discuss this methodology and the necessary tools for its implementation.
\cref{paper:olguinmunoz2018demoscaling} presents a short, high level overview of this methodological approach as applied to \gls{WCA}, making very straightforward assumptions about human behavior in \gls{WCA} applications.
\cref{paper:olguinmunoz2019edgedroid} extends the discussion from \cref{paper:olguinmunoz2018demoscaling}, providing more details on the implementation of the necessary measurement framework and the tooling employed.
This work presents the first empirical results obtained with our methodological approach, providing thus an initial validation of its utility for research.
\cref{paper:olguinmunoz2019edgedroid} also discusses in detail the assumptions that were made about human behavior and reactions.
We assume in these works a human which is impervious to poor system performance, and suffers no annoyance, fatigue, frustration, nausea or other shortcomings of real human users.
The result is a model of a user who responds in a precisely reproducible and deterministic manner to the same system stimulus every time.

\cref{paper:olguinmunoz2022cleave}, on the other hand, presents a discussion on the implementation of this methodology for \acp{NCS}.
As discussed above, these applications are similarly difficult as \gls{WCA} to benchmark (in particular at scale on multi-tenant systems) due to their client-side complexity, but in particular because of their often extreme sensitivity to latency.
\todo[inline]{More argumentation or refer to previous section.}
We apply thus in this work our methodology to \gls{NCS} through the implementation of a tool for the emulation and subsequent deployment of these systems on edge computing infrastructure.
This tool allows us to both emulate the physical components of a relatively simple control system plant and deploy real algorithms for its control.
The software acts as a middleware which abstracts away the network from the development of these workloads, allowing for quick prototyping and deployment.
This work also showcases the scalability and flexibility of our approach, allowing us to deploy scenarios with a large number of loops without the need for domain-specific hardware.

@@

Next, in \cref{paper:olguinmunoz2021impact,paper:olguinmunoz2023realistic} we delve deep into the implementation of our methodology for \gls{WCA}, as well as study its implications for optimization of these systems.
Although the above approach to human behavior in \gls{WCA} discussed represents a useful initial approximation, it is nonetheless not a realistic model of it.
In \cref{paper:olguinmunoz2021impact} we therefore take the first major step towards such a model, presenting a deep characterization of human behavior in these applications.
We develop this characterization through a human subject study with a cohort of \num{40} participants who were asked to interact with an instrumented \gls{WCA} application.
System responsiveness is altered in real-time during each execution of the task, and we record participants' reactions by measuring task- and system-related metrics, as well as biometrics from sensors placed on their bodies.
Participants were also asked to fill out two personality indicator questionnaires, allowing us to later correlate individual personality traits and measured reaction to changes in system responsiveness.

\cref{paper:olguinmunoz2023realistic} then concludes this line of work, building upon the insights and data obtained in \cref{paper:olguinmunoz2021impact} to develop the first ever data-driven model of human timings for \gls{WCA}.
The model is validated against previously obtained results, both through simulated, controlled executions and deployments on a real edge computing testbed.
This work also explores potential implications of this model for \gls{WCA} system optimization potential, particularly in the domains of energy consumption and sampling strategies.

### 2022-12-19 body/part1/overview/extending-wca.tex
glossaries

\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \ac{WCA}.
This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \ac{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.

We present in this paper the design and execution of a study in which undergraduate students of diverse fields of
study\footnote{%
    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \ac{CMU}.
}, were asked to interact with and follow the instructions given to them by a \ac{WCA} application.
Unbeknownst to the participants, the system responsiveness of the \ac{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.

@@

\begin{figure}[tb]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a}
        \caption{}\label{sfig:regularwcaexec}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b}
        \caption{}\label{sfig:delaywcaexec}
    \end{subfigure}
    \caption{%
        Comparison between normal task execution in the LEGO \ac{WCA}~\cite{chen2015early}~(\labelcref{sfig:regularwcaexec}) and modified task execution with delay in the experimental setup of \cref{paper:olguinmunoz2021impact}~(\labelcref{sfig:delaywcaexec}).
        In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target interval of time denoted \emph{delay}.
    }\label{fig:regularwca-vs-delaywca}
\end{figure}

@@

In order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
This array included devices to capture four physiological measures:
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item \ac{GSR} (also known as electrodermal activity)
    \item accelerometer data from the dominant wrist
    \item brain activity in the form of \ac{EEG}
    \item heart rate
\end{enumerate*}.
Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \ac{BFI} and the \ac{ITQ}.

@@

We introduce the first, to our knowledge, data-driven model for human timings in \ac{WCA} applications.
Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as
input past measurements of system responsiveness and produces realistic step execution times.
We also introduce a novel way to generate dynamic traces of frames for \ac{WCA} applications which can be combined
with the timing model for a full end-to-end emulation of a human.
We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid
approach~\cite{olguin2018scaling,olguin2019edgedroid}.

The study and benchmarking of \ac{WCA} applications is a challenging discipline due to these application's intrinsic
human-in-the-loop nature.
Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming
and prohibitively expensive for many research groups.

In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \ac{WCA},
the first data-driven model for human timings in \ac{WCA} applications.
This model represents a stochastic approach to execution time modeling which builds upon the data collected for
\textcite{olguinmunoz:impact2021}.
Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \ac{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.

>>>>

\cref{paper:olguinmunoz2021impact} describes a study designed to characterize the effects of system responsiveness on human behavior in step-based \gls{WCA}.
This corresponds to the first step in extending our EdgeDroid tool for the benchmarking of \gls{WCA}, first introduced in \cref{paper:olguinmunoz2018demoscaling,paper:olguinmunoz2019edgedroid}, with a more realistic model of human behavior.

We present in this paper the design and execution of a study in which undergraduate students of diverse fields of
study\footnote{%
    Participants were recruited from a pool of students enrolled in an undergraduate psychology course at \gls{CMU}.
}, were asked to interact with and follow the instructions given to them by a \gls{WCA} application.
Unbeknownst to the participants, the system responsiveness of the \gls{WCA} system was systematically altered in real time, and the resulting behavioral and physiological reactions were recorded.

@@

\begin{figure}[tb]
    \centering
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4a}
        \caption{}\label{sfig:regularwcaexec}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[t]{.49\textwidth}
        \centering
        \includegraphics[width=\textwidth]{publications/2021ImpactDelayedResponse/Fig4b}
        \caption{}\label{sfig:delaywcaexec}
    \end{subfigure}
    \caption{%
        Comparison between normal task execution in the LEGO \gls{WCA}~\cite{chen2015early}~(\labelcref{sfig:regularwcaexec}) and modified task execution with delay in the experimental setup of \cref{paper:olguinmunoz2021impact}~(\labelcref{sfig:delaywcaexec}).
        In the experimental task, an additional variable segment of time is introduced immediately following the processing of the input frame in order to extend the perceived processing time of the input to a specific target interval of time denoted \emph{delay}.
    }\label{fig:regularwca-vs-delaywca}
\end{figure}

@@

In order to capture metrics relating to the emotional response of humans to changes in system responsiveness, participants were asked to wear an array of biometric sensors.
This array included devices to capture four physiological measures:
\begin{enumerate*}[itemjoin={{, }}, itemjoin*={{, and }}]
    \item \gls{GSR} (also known as electrodermal activity)
    \item accelerometer data from the dominant wrist
    \item brain activity in the form of \gls{EEG}
    \item heart rate
\end{enumerate*}.
Additionally, to measure individual differences in personality between them, participants were asked to complete two questionnaires previous to beginning the task, the \gls{BFI} and the \gls{ITQ}.

@@

We introduce the first, to our knowledge, data-driven model for human timings in \gls{WCA} applications.
Using the data collected for \textcite{olguinmunoz:impact2021} as a base, we build a stochastic model which takes as
input past measurements of system responsiveness and produces realistic step execution times.
We also introduce a novel way to generate dynamic traces of frames for \gls{WCA} applications which can be combined
with the timing model for a full end-to-end emulation of a human.
We name this new model \emph{EdgeDroid 2.0}; a direct, more realistic evolution of our initial EdgeDroid
approach~\cite{olguin2018scaling,olguin2019edgedroid}.

The study and benchmarking of \gls{WCA} applications is a challenging discipline due to these application's intrinsic
human-in-the-loop nature.
Humans are notoriously unreliable, and greatly complicate the scalability and repeatability of experiments.
Furthermore, recruiting large enough cohorts of humans for large-scale experimentation is both greatly time-consuming
and prohibitively expensive for many research groups.

In the first half of this paper, we have introduced the EdgeDroid 2.0 model of human timing behavior for \gls{WCA},
the first data-driven model for human timings in \gls{WCA} applications.
This model represents a stochastic approach to execution time modeling which builds upon the data collected for
\textcite{olguinmunoz:impact2021}.
Together with this model, we have also introduced a novel procedure for the generation of synthetic traces of frames in step-based \gls{WCA}, allowing for a full end-to-end emulation of a human when combined with the timing model.

### 2022-12-19 body/part1/overview/implications-optimization.tex
glossaries

\section{Implications for the optimization of \ac{WCA}}

\begin{enumerate}
    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \ac{WCA}.
    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \ac{WCA} when considering human behavior using our model.
    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \ac{WCA} optimization; number of processed samples and energy consumption per step.
    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
\end{enumerate}

Next, we have explored the impact of such a realistic model on the application lifetime footprint of \ac{WCA} applications.
We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.

Finally, we have delved into the potential for optimization in \ac{WCA} systems using the previously discussed timing models.
We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \ac{WCA} which results in an adaptive sampling strategy.
We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \ac{WCA} applications.

>>>>

\section{Implications for the optimization of \gls{WCA}}

\begin{enumerate}
    \item\label{item:contrib:footprint} Using the model described above, we study the implications of realistic human behavior on the application lifetime footprint of \gls{WCA}.
    In concordance with previous work~\cite{olguinmunoz:impact2021}, we find dependencies between system responsiveness and human step execution times that lead to substantially different application lifetimes when compared to a first-order baseline which does not take into account human behavior.
    \item\label{item:contrib:optimization} Finally, we study the potential for optimization in \gls{WCA} when considering human behavior using our model.
    We develop a generic model for the stochastic optimization of resource consumption versus responsiveness trade-offs in these applications, which we apply to two potential avenues for \gls{WCA} optimization; number of processed samples and energy consumption per step.
    Compared to the state-of-the-art, our optimization model results in up to \textasciitilde\SI{60}{\percent} fewer samples processed or an average of \SI{20}{\percent} less energy consumed per step, while maintaining comparable levels of responsiveness.
\end{enumerate}

Next, we have explored the impact of such a realistic model on the application lifetime footprint of \gls{WCA} applications.
We have shown that less realistic modeling approaches which do not take into account higher-order effects on execution time distributions can potentially lead to significant misestimations of application footprint.

Finally, we have delved into the potential for optimization in \gls{WCA} systems using the previously discussed timing models.
We have proposed a novel stochastic optimization framework for resource consumption-system responsiveness trade-offs in \gls{WCA} which results in an adaptive sampling strategy.
We have shown that this framework is applicable to a myriad of metrics in these applications, and showcased experimental results employing this framework for the minimization of number of samples processed per step and total energy consumption per step.
Our results show up to a \SI{50}{\percent} increase in performance with respect to state of the art when optimizing for number of samples, and up to a \SI{30}{\percent} improvement when optimizing for energy consumption, proving thus the value of such frameworks for the design of \gls{WCA} applications.
